### 2014-11-07 concl.tex
edits

In this paper, we have presented \name a shallow Haskell extension for
declarative programming over eventually consistent data stores. The key idea of
\name is the automatic classification of fine-grained consistency contracts on
operations and distributed transactions with respect to the consistency and
isolation levels offered by the store. Our contract language is carefully
crafted from a decidable subset of first-order logic, enabling the use of
automatic theorem prover to discharge the proof obligations associated with
contract classification. We provide meta-theory that certifies the correctness
of contract classification. We realize an instantiation of \name on top of
off-the-shelf distributed store, Cassandra, and illustrate the benefit of
fine-grained contract classification by implementing and evaluating several
scalable applications.

>>>>

This paper presents \name, a shallow Haskell extension for declarative
programming over eventually consistent data stores. The key idea underlying
\name's design is the automatic classification of fine-grained consistency
contracts on operations and distributed transactions with respect to the
consistency and isolation levels offered by the store. Our contract language
is carefully crafted from a decidable subset of first-order logic, enabling
the use of automated verification tools to discharge the proof obligations
associated with contract classification. We realize an instantiation of
\name on top of off-the-shelf distributed store, Cassandra, and illustrate
the benefit of fine-grained contract classification by implementing and
evaluating several scalable applications.

### 2014-11-07 evaluation.tex
edits

In this section, we evaluate \name programs, report their contract profile and
illustrate the performance benefits of fine-grained consistency classification
on operations and transactions. We also evaluate on the impact of the
summarization. We implemented the following applications, which includes
individual RDTs as well as larger applications composed of several RDTs:

\begin{itemize}
\setlength{\itemsep}{2pt}
\item \textbf{LWW register}: A last-write-wins register that provides read and
write operations. Each write is associated with a timestamp, which is used to
resolve conflicting concurrent writes -- newer write wins.

\item \textbf{DynamoDB register}: A integer register that allows eventual and
strong puts and gets, conditional puts, increment and decrement operations.

\item \textbf{Bank account}: Our running example, with savings and current
accounts.

\item \textbf{Shopping list}: Collaborative shopping list which allows adding
and deleting items.

\item \textbf{Online store}: Models an online store with shopping cart and
dynamically changing item prices. Checkout process verifies that the customer
only pays the accepted price.

@@

\item \textbf{Microblog}: A twitter-like microblogging site, modelled after
Twissandra~\cite{Twissandra}. The application allows adding a new user, adding
and replying to tweets, following, unfollowing and blocking users, and fetching
a user's timeline, userline, followers and following.
\end{itemize}

@@

For our performance evaluation, we deploy \name applications in \emph{clusters},
where each cluster is composed of 5 fully replicated Cassandra replicas within
the same datacenter. We instantiate one shim layer node for every Cassandra
replica, and place it on the same VM as the Cassandra replica. Clients are
instantiated on the same data center as the store, and run transactions. We
deploy the each node in the cluster on \cf{c3.4xlarge} Amazon EC2 instances.
Our shim layer nodes are multi-threaded, and we allocate 8 CPUs (out of 16
available) for each shim layer node. The clients also run on \cf{c2.4xlarge}
instances. We call this \cf{1DC} configuration. For our geo-distributed
experiments (\cf{2DC}), we instantiate 2 clusters, each with 5 nodes, and place
the clusters on US-east (Virginia) and US-west (Oregon). The average
inter-region latency was 85ms.

@@

Figure~\ref{grf:BA-tp-vs-lat} shows the throughput vs. latency of operations
in bank account example as we increase the number of clients in \cf{1DC}
configuration. Our client workload was generated using YCSB
benchmark~\cite{YCSB}. The benchmark uniformly chose from 100,000 keys, where
the operation spread was 25\% withdraw, 25\% deposit and 50\% getBalance, which
corresponds to the default 50:50 read:write mix in YCSB. We increased the
number of clients from 128 to 1024, and each experiment ran for 180 seconds.

The lines marked EC and CC correspond to all operations being assigned EC and
CC consistency levels. These levels compromise correctness as withdraw has to
be and SC operation. The line SC corresponds to a configuration where all
operations are strongly consistent; this ensures application correctness. \name
corresponds to our implementation, which classifies operations based on their
contracts. Both \name and SC ensure correctness. However, with 512 clients,
\name implementation was within 41\% of latency and 18\% of throughput of EC,
whereas SC operations had 162\% higher latency and 52\% lower throughput than
EC operations. Observe that there is a point in each case after which the
latency increases while the throughput decreases. This indicates a point after
which the store becomes saturated with client requests.

In \cf{2DC} configuration (not-shown), the average latency of SC operations
with 512 clients increased by 9.4$\times$ due to the cost of geo-distributed
coordination, whereas \name operations were only 2.2$\times$ slower, mainly due
to the increased cost of \cf{withdraw} operations. Importantly, the latency of
\cf{getBalance} and \cf{deposit} remained almost the same. This illustrates the
benefit of fine-grained contract classification in \name.

We compare the performance of different transaction isolation level choices in
Figure~\ref{grf:LWW-txn} using the LWW register. The numbers were obtained
under 1DC configuration. The YCSB workload was modified to issue 10 operations
per transaction, with the default 50:50 read:write mix. Each operation is
assumed to have eventual consistency. \cf{NoTxn} corresponds to a configuration
that does not use transactions. Compared to this RC is only 12\% shower in
terms of latency with 512 clients, where as RR is 2.3X slower. The difference
between RC and \cf{NoTxn} is due to the meta-data overhead of recording
transaction information in the object state. For RR transaction, the cost of
capturing and maintaining the snapshot in an RR transaction is the biggest
source of overhead.

We also compared (not shown) the performance of EC LWW operations directly
against Cassandra (our backing store), which uses last-writer-wins as the only
convergence semantics. While Cassandra provides no stronger-than-eventual
consistency properties, \name was within 30\%(20\%) of latency(throughput) of
Cassandra with 512 clients, illustrating that the programmers only have to pay
a minimal overhead for the expressive and stronger \name programming model.

Figure~\ref{grf:rubis} compares the \name implementation of RUBiS in \cf{1DC}
configuration against a single replica (NoRep) and strongly replicated
(StrongRep) \cf{1DC} deployment. The benchmark was RUBiS bidding mix, which has
15\% read-write interactions, which is representative of the auction workload.
Without replication, NoRep trivially provides strong consistency. However, this
deployment does not scale beyond 1750 operations per second. Strong replication
offers better throughput at the cost of greater latency due to inter-replica
coordination. \name deployment offers the benefit of replication, while only
paying the cost of coordination when necessary.

Finally, we study the impact of summarization in
Figure~\ref{grf:summarization}. We utilize 128 clients and a single \name
replica, with all the clients operating on the \emph{same} LWW register to
stress test the summarization mechanism. The shim layer cache (mem) of
operations is summarized every 64 updates, while the updates in the backing
store (disk) are summarized every 4096 updates. Each point in the graph
represents the average latency of the previous 1000 operations. Each experiment
is run for 60s.

The results show that without summarization, the average latency of operations
increase exponentially to almost 1 second, and only 13K operations were
performed in a minute. Since every operation has to reduce over the set of all
previous operations, with a ever growing set, the operations take increasingly
more time to complete. With summarization only in memory, the performance still
degrades due to the cost of fetching all previous updates from the backing
store into the shim layer. Fetching the latest updates is essential for SC
operations. With both summarizations enabled, we see that the latency does not
increase over time, and we were able to perform 67K operations. This graph
illustrates the importance and effectiveness of summarization in \name.

>>>>

We present an evaluation study of our implementation, report contract
profiles of benchmark programs, and illustrate the performance benefits of
fine-grained consistency classification on operations and transactions. We
also evaluate the impact of the summarization. We have implemented the
following applications, which includes individual RDTs as well as larger
applications composed of several RDTs:

\begin{itemize}
\setlength{\itemsep}{2pt}
\item \textbf{LWW register}: A last-write-wins register that provides read
  and write operations. Each write is associated with a timestamp, which is
  used to resolve conflicting concurrent writes -- a newer write always
  wins.

\item \textbf{DynamoDB register}: An integer register that allows eventual
  and strong puts and gets, conditional puts, increment and decrement
  operations.

\item \textbf{Bank account}: Our running example, with savings and checking
  accounts.

\item \textbf{Shopping list}: A collaborative shopping list that allows
  adding and deleting items.

\item \textbf{Online store}: An online store with shopping cart
  functionality and dynamically changing item prices.  The checkout process
  verifies that the customer only pays the accepted price.

@@

\item \textbf{Microblog}: A twitter-like microblogging site, modelled after
  Twissandra~\cite{Twissandra}. The application allows adding new users,
  adding and replying to tweets, following, unfollowing and blocking users,
  and fetching a user's timeline, userline, followers and following.
\end{itemize}

@@

For our performance evaluation, we deploy \name applications in
\emph{clusters}, where each cluster is composed of five fully replicated
Cassandra replicas within the same datacenter. We instantiate one shim layer
node co-located on the same VM as a Cassandra replica. Clients are
instantiated within the same data center as the store, and run
transactions. We deploy each node in the cluster on \cf{c3.4xlarge} Amazon
EC2 instances.  Our shim layer nodes are multi-threaded, and we allocate 8
CPUs (out of 16 available) for each shim layer node. The clients also run on
\cf{c2.4xlarge} instances. We call this a \cf{1DC} configuration. For our
geo-distributed experiments (\cf{2DC}), we instantiate 2 clusters, each with
five nodes, and place the clusters on US-east (Virginia) and US-west
(Oregon) locations. The average inter-region latency was 85ms.

@@

Figure~\ref{grf:BA-tp-vs-lat} shows throughput vs. latency of operations
in the bank account example as we increase the number of clients in a \cf{1DC}
configuration. Our client workload was generated using the YCSB
benchmark~\cite{YCSB}. The benchmark uniformly chooses from 100,000 keys, where
the operation spread was 25\% withdraw, 25\% deposit and 50\% getBalance, which
corresponds to the default 50:50 read:write mix in YCSB. We increased the
number of clients from 128 to 1024, and each experiment ran for 180 seconds.

The lines marked EC and CC correspond to all operations (including
\cf{withdraw}) being assigned EC and CC consistency levels. These levels
compromise correctness as \cf{withdraw} has to be an SC operation. The SC
line corresponds to a configuration where all operations are strongly
consistent; this ensures application correctness, at the cost of
performance. \name corresponds to our implementation, which classifies
operations based on their contract specifications.  Both \name and SC ensure
correctness. With 512 clients, the \name implementation was within 41\% of
the latency and 18\% of the throughput of EC, whereas SC operations had
162\% higher latency and 52\% lower throughput than EC operations. Observe
that there is a point in each case after which the latency increases while
the throughput decreases; these points correspond to client request
saturation points in the data store.

In a \cf{2DC} configuration (not shown here), the average latency of SC
operations with 512 clients increased by 9.4$\times$ due to the cost of
geo-distributed coordination, whereas \name operations were only 2.2$\times$
slower, mainly due to the increased cost of \cf{withdraw}
operations. Importantly, the latency of \cf{getBalance} and \cf{deposit}
remained almost the same, illustrating the benefit of fine-grained
contract classification.

We compare the performance of different transaction isolation level choices
in Figure~\ref{grf:LWW-txn} using the LWW register. The numbers were
obtained under a 1DC configuration. The YCSB workload was modified to issue
10 operations per transaction, with a default 50:50 read:write mix. Each
operation is assumed to be eventual consistent. \cf{NoTxn} corresponds to a
configuration that does not use transactions. Compared to this, RC is only
12\% shower in terms of latency with 512 clients, whereas RR is 2.3X
slower. The difference between RC and \cf{NoTxn} is due to the meta-data
overhead of recording transaction information in the object state. For RR
transactions, the cost of capturing and maintaining a snapshot is the
biggest source of overhead.

We also compared (not shown) the performance of EC LWW operations directly
against Cassandra, which uses last-writer-wins as the only convergence
semantics. While Cassandra provides no stronger-than-eventual consistency
properties, \name was within 30\%(20\%) of latency(throughput) of Cassandra
with 512 clients, supporting our thesis that programmers only have to incur
relatively low overhead for a more expressive programming model which
provides stronger provable consistency guarantees.

Figure~\ref{grf:rubis} compares the \name implementation of RUBiS in a
\cf{1DC} configuration against a single replica ({\sf NoRep}) and a strongly
replicated ({\sf StrongRep}) \cf{1DC} deployment. The benchmark uses the
default RUBiS bidding mix, which has 15\% read-write interactions, which is
representative of the auction workload.  Without replication, {\sf NoRep}
trivially provides strong consistency. However, this deployment does not
scale beyond 1750 operations per second. Strong replication offers better
throughput at the cost of greater latency due to inter-replica
coordination. The \name deployment offers the benefit of replication, while
only paying the cost of coordination when necessary.

Finally, we study the impact of summarization in
Figure~\ref{grf:summarization}. We use 128 clients and a single \name
replica, with all clients operating on the \emph{same} LWW register to
stress test the summarization mechanism. The shim layer cache (memory) is
summarized every 64 updates, while the updates in the backing store (disk)
are summarized every 4096 updates. Each point in the graph represents the
average latency of the previous 1000 operations. Each experiment is run for
one minute.  Without summarization, the average latency of operations
increases exponentially to almost one second, and only 13K operations were
performed in a minute. Since every operation has to reduce over the set of
all previous operations, operations take increasingly more time to complete
since they must contend with an ever growing set.  With summarization only
in memory, performance still degrades due to the cost of fetching all
previous updates from the backing store into the shim layer. Fetching the
latest updates is essential for SC operations. With summarization enabled on
both disk and memory, latency does not increase over time, and the
implementation realizes throughput of 67K operations/minute.

### 2014-11-07 implementation.tex
edits

Replicated data types and the stronger consistency semantics are implemented
and enforced in the \emph{shim layer}. Our implementation supports eventual,
causal, and strong consistency for data type operations, and RC, MAV, and RR
semantics for transactions.  This functionality is implemented entirely on top
of the standard interface exposed by Cassandra. From an engineering
perspective, leveraging an off-the-shelf data store enables an implementation
comprising roughly only 2500 lines of Haskell code, which is packaged as a
library.

@@

Each new effect in \name is realized as a row insertion in Cassandra, and the
state of an object is the set of all corresponding rows. The shim layer
maintains a causally consistent in-memory snapshot of a subset of objects in
the backing store, by explicitly tracking dependencies introduced between
effects due to visibility, session and same transaction relations. Dependence
tracking is similar to the techniques presented in~\cite{BoltOn}
and~\cite{Eiger}. Because Cassandra provides durability, convergence and fault
tolerance, each shim layer node simply acts as a soft-state cache, do not
communicate directly with each other, and can safely be terminated at any
point. Similarly, new shim layer nodes can be spawned on demand.

The shim layer nodes periodically fetch updates from the backing store for
eventually consistent operations, and on-demand for causally consistent and
strongly consistent operations. Strongly consistent operations are performed
after obtaining exclusive leases on objects. The lease mechanism is implemented
with the help of Cassandra's support for conditional updates and expiring
columns. Cassandra does not provide general-purpose transactions. Since the
transaction guarantees provided by \name are
coordination-free~\cite{BailisHAT}, we realize efficient implementations by
explicitly tracking the dependencies between operations and transactions.
Importantly, the weaker isolation semantics of transactions in \name permit
transactions to be discharged if at least one shim layer node is reachable.

@@

The main challenge in realizing an efficient implementation of operation-based
replicated data types is that the state of the object defined in terms of its
set of effects grows with every effectful operation. If left unchecked,
performance slow down over time, until the shim layer memory or backing store
disk runs out of memory. Luckily, the state of the operation-based replicated
data type can often be summarized to an \emph{observably equivalent} smaller
state. For example, a last-writer-wins register with multiple updates can be
summarized to an equivalent register with only the last write. Similarly, a
bank account with a series of deposits and updates with balance $b$ is
equivalent to a bank account with a single deposit of $b$. A set with a
collection of add and remove operations is equivalent to a set with a series of
addition of live elements from the original set.

@@

Performing summarization in the backing store is more complicated; the whole
process needs to be atomic from a client's perspective, but Cassandra does not
provide multi-row transactions. We have engineered and implemented a scalable
summarization mechanism for the backing that permits concurrent client
operations, but nonetheless prohibits these operations from witnessing
intermediate states of the summarization process.

>>>>

Replicated data types and various consistency semantics are implemented and
enforced in the \emph{shim layer}. Our implementation supports eventual,
causal, and strong consistency for data type operations, and RC, MAV, and RR
semantics for transactions.  This functionality is implemented entirely on
top of the standard interface exposed by Cassandra. From an engineering
perspective, leveraging an off-the-shelf data store enables an
implementation comprising roughly only 2500 lines of Haskell code, which is
packaged as a library.

@@

Each new effect in \name is realized as a row insertion in Cassandra, and
the state of an object is the set of all corresponding rows. The shim layer
maintains a causally consistent in-memory snapshot of a subset of objects in
the backing store, by explicitly tracking dependencies introduced between
effects due to visibility, session and same transaction
relations. Dependence tracking is similar to the techniques presented
in~\cite{BoltOn} and~\cite{Eiger}. Because Cassandra provides durability,
convergence, and fault tolerance, each shim layer node simply acts as a
soft-state cache, with no inter-node communication, and can safely be
terminated at any point. Similarly, new shim layer nodes can be spawned on
demand.

The shim layer nodes periodically fetch updates from the backing store for
eventually consistent operations, and on-demand for causally consistent and
strongly consistent operations. Strongly consistent operations are performed
after obtaining exclusive leases on objects. The lease mechanism is
implemented with the help of Cassandra's support for conditional updates and
expiring columns. Cassandra does not provide general-purpose
transactions. Since the transaction guarantees provided by \name are
coordination-free~\cite{BailisHAT}, we realize efficient implementations by
explicitly tracking dependencies between operations and transactions.
Importantly, the weaker isolation semantics of transactions in \name permit
transactions to be discharged if at least one shim layer node is reachable.

@@

The main challenge in realizing an efficient implementation of operation-based
replicated data types is that the state of the object defined in terms of its
set of effects grows with every effectful operation. If left unchecked,
performance slows down over time, until the shim layer memory or backing store
disk runs out of memory. Luckily, the state of the operation-based replicated
data type can often be summarized to an \emph{observably equivalent} smaller
state. For example, a last-writer-wins register with multiple updates can be
summarized to an equivalent register with only the last write. Similarly, a
bank account with a series of deposits and updates with balance $b$ is
equivalent to a bank account with a single deposit of $b$. A set with a
collection of add and remove operations is equivalent to a set with a series of
addition of live elements from the original set.

@@

Performing summarization in the backing store is more complicated since the
whole process needs to be atomic from a client's perspective, but Cassandra
does not provide multi-row transactions. We have engineered and implemented
a scalable summarization mechanism for the backing store that permits
concurrent client operations, but nonetheless prohibits these operations
from witnessing intermediate states of the summarization process.
